---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


# AfterWork Data Science: Data Visualisation with Python Project

## Install packages
```{r}
install.packages("Tidyverse")
install.packages("reshape2")
```

```{r}
library(tidyverse)
library(reshape2)
```

```{r}
library(reprex)
```


## 1. Defining the Question

### a) Specifying the Data Analysis Question
dataset contains information on the various chain of hotels, including when customers made
the booking, length of stay, the number of adults, children, or babies, and the number of available parking spaces, among other things.
.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
The goal is to help the management decide on the strategies that will lead to revenue growth. We will focus more 


### d) Recording the Experimental Design
 We will perform data exploration, data cleaning, and analysis to come with appropriate
recommendations. we must have a visualization for each question.


### e) Data Relevance
The data contains all the fields required to answer the research question.


## 2. Reading the Data

```{r}
# Load the data below
# --- 
# Dataset url = https://bit.ly/2WvQbhJ
# --- 
# YOUR CODE GOES BELOW
# 
hotel_booking_df <- read.csv("hotel_bookings.csv", header = TRUE, na.strings =c("NULL", NA))

```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(hotel_booking_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(hotel_booking_df, 5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample_n(hotel_booking_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(hotel_booking_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
glimpse(hotel_booking_df)
```

```{r}
unique(hotel_booking_df$adr)
```

##Record your general observations below:

The data set has 119390 rows and 32 columns
All columns have the correct data types

## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
str(hotel_booking_df)

sum(is.na(hotel_booking_df))
```

#### We observe the following from our dataset:

Observation 1: There are 478 missing values in our data set

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
# hotel_booking_df[duplicated(hotel_booking_df), ]
sum(duplicated(hotel_booking_df))
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 
which(colSums(is.na(hotel_booking_df))  == ncol(hotel_booking_df))
```

We observe the following from our dataset:

Observation 1: There are no columns in which all the data is null

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW

which(rowSums(is.na(hotel_booking_df)) == nrow(hotel_booking_df))
```

We observe the following from our dataset:

Observation 1:  there is no row in which all the data is null

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 

head(hotel_booking_df)

# Select the desired columns
hotel_booking_df2 <- hotel_booking_df %>%
  na.omit() %>%
  select(is_canceled, stays_in_weekend_nights, stays_in_week_nights, adults, children, babies, is_repeated_guest, previous_cancellations, previous_bookings_not_canceled, days_in_waiting_list, total_of_special_requests)
hotel_booking_df2

# Create a correlation matrix
corr_matrix <- cor(hotel_booking_df2, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df

# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "black") +
  geom_text(aes(label = round(value, 1))) +
  coord_fixed() +
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_distiller(palette = "RdPu") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(size=9, angle=45, hjust = 1),
    axis.title.y = element_blank()
  )

```

We observe the following from our data set:

Observation 1: There's a high correlation between "previous booking cancellation" and "is repeated guests" therefore Guests who have not cancelled their previous bookings are more likely to visit the hotel in future.

Observation 2: There's a negative correlation between cancellation and special requests meaning customers with more special requests are less likely to cancel their bookings

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
hotel_booking_df = select(hotel_booking_df, -company)
head(hotel_booking_df)
```

From the data variable description we see that the Distribution Channel category that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#

hotel_booking_df$agent[is.na(hotel_booking_df$agent)] <- "TO"
hotel_booking_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#

hotel_booking_df <- subset(hotel_booking_df, adults!=0 | children!=0 | babies!=0)
hotel_booking_df
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
hotel_booking_df$children[is.na(hotel_booking_df$children)] <- round(mean(hotel_booking_df$children, na.rm = TRUE), 0)
hotel_booking_df
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
sum(is.na(hotel_booking_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
# hotel_booking_df$children <- lapply(hotel_booking_df$children, as.numeric)
as.integer(hotel_booking_df$children)
hotel_booking_df
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot

hotel_booking_df %>%
  group_by(reservation_status) %>%
  summarise(is_canceled = sum(is_canceled))

ggplot(hotel_booking_df, aes(x= reservation_status, y= is_canceled))+
  geom_col(stat = "identity", width =  0.5, fill= "blue") +
  labs(
    title = "Reservation Status",
    x= "Status",
    y= "Number"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Ariel Black", color="purple")
  )
  
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

resort_city_hotel <- hotel_booking_df %>%
  count(hotel)
resort_city_hotel

ggplot(data=resort_city_hotel, aes(x="", y=n, fill=hotel)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") +
  geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette="Set2")+
  labs(
    fill = "Hotel",
    title = "Booking Between Resort Hotel and City Hotel"
  )+
  theme_void()
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# 

year_booking <- hotel_booking_df %>%
  count(arrival_date_year)
year_booking

ggplot(data=year_booking, aes(x="", y=n, fill=factor(arrival_date_year))) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") +
    geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set2") +
  labs(
    fill = "Years",
    title = "Booking for each year") +
  theme_void()
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 

#Arrange data in order
busy_months <- hotel_booking_df %>%
  mutate(arrival_date_month = factor(arrival_date_month, levels = c("January", "February","March", "April","May","June","July", "August","September","October","November","December")))%>%
 count(arrival_date_month)
busy_months

ggplot(busy_months, aes(x= arrival_date_month, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "sky blue") +
  labs(
    title = "Bookings per Month",
    x= "Month",
    y= "Number of bookings",
  ) +
  theme(
    plot.title = element_text(hjust = 0.5,  size=16, family="Arial", color="black")
  )
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#

hotel_booking_df

countries <- hotel_booking_df %>%
  group_by(country) %>%
  summarise(guests = sum(adults)+sum(children)+sum(babies)) %>%
  # arrange(desc(guests)) %>%
  top_n(3)
countries

ggplot(countries, aes(x= country, y= guests))+
  geom_col(stat = "identity", width =  0.8, fill= "blue") +
  labs(
    title = "Top 3 Countries per Guests",
    x= "Countries",
    y= "Number of Guests"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="black")
  )+
  coord_flip()
```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 

stay_length <- hotel_booking_df %>%
  group_by(hotel) %>%
  summarise(total_stay_days = sum(stays_in_weekend_nights)+sum(stays_in_week_nights))
stay_length

ggplot(data = stay_length, aes(x= hotel, y= total_stay_days))+
  geom_col(stat = "identity", width =  0.8, fill= "grey") +
  labs(
    title = "Days Stayed per Hotel",
    x= "Hotel Type",
    y= "Number of Days Stayed"
  ) +
  theme(
    plot.title = element_text(hjust = 0, size=16, family="Arial", color="blue")
  )+
  theme_classic()
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
accommodation <- hotel_booking_df %>%
  count(reserved_room_type)
accommodation

ggplot(data = accommodation, aes(x= reserved_room_type, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "brown") +
  labs(
    title = "Bookings per Accomodation Type",
    x= "Room Type",
    y= "Number of Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )+
coord_flip()
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:

1. PRT region had the highest number of guests while France had the lowest number of guests
2. Room Type A had the highest number of bookings while room type L had the least number of bookings
3. Most guests prefer city hotels to resorts


## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.


When should hotels provide special offers?

Hotels should provide offers between November and January because that's when they had the least number of visitors 



## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?
Yes, though it would have been important to get customer rating and Total charge 

b). Do we need other data to answer our question?
It would have been important to view the amount paid per room since the goal was to improve revenue.

c). Did we have the right question?

Questions asked were relevant


